---
title: "OpenFDA Medical Device Adverse Event Analysis"
subtitle: "Analysis of FDA Medical Device Adverse Event Reports"
author: "Generated by OpenFDA Analysis Pipeline"
date: today
lof: true
lot: true
toc: true
version: 1.0
format:
  medstata-typst:
    default-image-extension: svg
execute:
  echo: false
  warning: false
  message: false
  fig-width: 7
  fig-height: 5
  freeze: auto
---

```{r setup}
#| echo: false
#| results: 'hide'
#| message: false

# Load required libraries
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)

# Custom ggplot2 theme matching base R aesthetics
theme_baseR <- function(base_size = 12) {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.grid = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
      axis.line = element_blank(),
      axis.ticks = element_line(color = "black"),
      plot.title = element_text(
        hjust = 0.5,
        face = "bold",
        margin = margin(b = 20, t = 10)
      ),
      plot.margin = unit(c(1.5, 1.5, 1, 1.5), "lines"),
      plot.title.position = "plot"
    )
}
theme_set(theme_baseR())

# Save current locale and set to English
original_locale <- Sys.getlocale("LC_TIME")
Sys.setlocale("LC_TIME", "C")

# Color palette using brewer colors
brbg_colors <- scales::brewer_pal(palette = "BrBG")(6)
colors <- list(
  primary   = brbg_colors[1],
  secondary = brbg_colors[2],
  accent    = brbg_colors[3],
  highlight = brbg_colors[4],
  warning   = brbg_colors[5],
  info      = brbg_colors[6]
)

# Configuration
PATIENT_EXCLUSIONS <- c(
  "No Code Available",
  "No Known Impact Or Consequence To Patient",
  "Symptoms or Conditions",
  "No Information",
  "No Consequences Or Impact To Patient",
  "Appropriate Clinical Signs",
  "No Clinical Signs",
  "Conditions Term / Code Not Available",
  "Insufficient Information",
  "No Patient Involvement",
  "Reaction",
  "Patient Problem/Medical Problem"
)

PRODUCT_EXCLUSIONS <- c(
  "Adverse Event Without Identified Device or Use Problem",
  "Appropriate Term/Code Not Available",
  "Unknown (for use when the device problem is not known)",
  "Insufficient Information",
  "No Apparent Adverse Event"
)

# Helper functions
parse_problem_lists <- function(series) {
  series %>%
    str_remove_all("\\[|\\]|'") %>%
    str_split(", ") %>%
    unlist()
}

shorten_brand_name <- function(brand, max_words = 4) {
  words <- str_split(brand, " ")[[1]]
  if (length(words) > max_words) {
    paste(words[1:max_words], collapse = " ")
  } else {
    brand
  }
}
```

```{r load-data}
#| echo: false

# Read the data
data <- read_csv("processed_data.csv", show_col_types = FALSE)
```

```{r compute-summary-stats}
#| echo: false

# Compute all summary statistics
total_reports <- nrow(data)
date_min <- min(data$date_received, na.rm = TRUE)
date_max <- max(data$date_received, na.rm = TRUE)
duration_days <- as.numeric(date_max - date_min)
duration_months <- round(duration_days / 30.44, 1)

unique_manufacturers <- n_distinct(data$manufacturer_std)
unique_brands <- n_distinct(data$brand_std)

# Patient problems statistics
patient_problems_all_list <- parse_problem_lists(data$patient_problems)

reports_with_patient_problems <- sum(
  !patient_problems_all_list %in% PATIENT_EXCLUSIONS
)

# Monthly statistics
monthly_counts <- data %>%
  mutate(year_month = floor_date(date_received, "month")) %>%
  count(year_month) %>%
  arrange(year_month)

avg_monthly_reports <- round(mean(monthly_counts$n), 1)
max_monthly_reports <- max(monthly_counts$n)
max_month <- monthly_counts %>%
  filter(n == max_monthly_reports) %>%
  pull(year_month) %>%
  format("%B %Y")
```

# Executive Summary

This report analyzes **`r format(total_reports, big.mark=",")`** FDA medical device adverse event reports submitted between **`r format(date_min, "%B %d, %Y")`** and **`r format(date_max, "%B %d, %Y")`** (a period of **`r duration_months`** months). The dataset includes **`r format(unique_manufacturers, big.mark=",")`** unique manufacturers and **`r format(unique_brands, big.mark=",")`** unique device brands. The average reporting rate was **`r avg_monthly_reports`** reports per month, with peak reporting of **`r max_monthly_reports`** reports in **`r max_month`**.

```{r summary-table}
#| label: tbl-summary-stats
#| tbl-cap: "Summary statistics of FDA MAUDE reports"

summary_stats <- tibble(
  Metric = c(
    "Total Reports",
    "Date Range",
    "Reporting Duration",
    "Unique Manufacturers",
    "Unique Device Brands",
    "Average Monthly Reports",
    "Maximum Monthly Reports"
  ),
  Value = c(
    format(total_reports, big.mark = ","),
    paste(format(date_min, "%B %d, %Y"), "to", format(date_max, "%B %d, %Y")),
    paste(duration_days, "days (", duration_months, "months)"),
    format(unique_manufacturers, big.mark = ","),
    format(unique_brands, big.mark = ","),
    paste(avg_monthly_reports, "reports/month"),
    paste(max_monthly_reports, "reports in", max_month)
  )
)

kable(summary_stats, align = c("l", "l"))
```

# Methodology

## Data Source

The search of the FDA Manufacturer and User Facility Device Experience (MAUDE) database was conducted using the openFDA Device Event API. Retrieved JSON-formatted data were downloaded directly from the API and subsequently processed and cleaned for analysis.

## Data Standardization

The analysis employs fuzzy matching algorithms to standardize manufacturer and brand names, addressing inconsistencies in naming conventions across reports. This standardization process uses the **RapidFuzz** library with partial ratio matching to group similar names under a canonical representation.

## Problem Classification

Patient and product problems are extracted and categorized, excluding non-informative categories to focus on meaningful data patterns.

# Temporal Trend Analysis

## Overall Reporting Trends

```{r temporal-analysis}
#| label: fig-temporal-trends
#| fig-cap: "Monthly trend of FDA MAUDE reports"
#| fig-width: 8
#| fig-height: 6

monthly_plot_data <- data %>%
  mutate(year_month = floor_date(date_received, "month")) %>%
  count(year_month)

# Calculate appropriate date breaks based on data range
date_range_years <- as.numeric(difftime(max(data$date_received), 
                                        min(data$date_received), 
                                        units = "days")) / 365.25

if(date_range_years > 2) {
  date_labels <- "%Y"
  date_breaks <- "1 year"
} else {
  date_labels <- "%b %Y"
  date_breaks <- "3 months"
}

ggplot(monthly_plot_data, aes(x = year_month, y = n)) +
  geom_line(color = colors$primary, linewidth = 1.2) +
  geom_point(color = colors$primary, size = 3) +
  geom_smooth(
    method = "loess", 
    se = TRUE, 
    color = colors$secondary, 
    fill = colors$accent, 
    alpha = 0.2
  ) +
  scale_x_date(date_breaks = date_breaks, date_labels = date_labels) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Monthly FDA MAUDE Reports",
    subtitle = paste(format(date_min, "%B %Y"), "-", format(date_max, "%B %Y")),
    x = "Month",
    y = "Number of Reports",
    caption = "Source: FDA MAUDE Database"
  )
```

```{r compute-trend-stats}
#| echo: false

# Identify peaks (above 75th percentile)
threshold <- quantile(monthly_plot_data$n, 0.75)
peak_months <- monthly_plot_data %>%
  filter(n > threshold) %>%
  arrange(desc(n)) %>%
  mutate(month_label = format(year_month, "%B %Y"))

peak_list <- paste(peak_months$month_label, collapse = ", ")
sd_monthly <- round(sd(monthly_plot_data$n), 1)
```

::: {.callout-note}
## Trend Observation
The reporting shows variability across months, with notable peaks in **`r peak_list`**. The average monthly reporting rate is **`r avg_monthly_reports`** reports, with a standard deviation of **`r sd_monthly`** reports.
:::

## Cumulative Reports Over Time

```{=typst}
#set page(
  flipped: true,
)
```

```{r cumulative-plot}
#| label: fig-cumulative-reports
#| fig-cap: "Cumulative FDA MAUDE reports over time"
#| fig-width: 10
#| fig-height: 6

cumulative_data <- data %>%
  arrange(date_received) %>%
  mutate(cumulative_count = row_number())

# Calculate appropriate date breaks
date_range_years <- as.numeric(difftime(max(data$date_received), 
                                        min(data$date_received), 
                                        units = "days")) / 365.25

if(date_range_years > 2) {
  date_labels <- "%Y"
  date_breaks <- "1 year"
} else {
  date_labels <- "%b %Y"
  date_breaks <- "3 months"
}

ggplot(cumulative_data, aes(x = date_received, y = cumulative_count)) +
  geom_line(color = colors$warning, linewidth = 1) +
  geom_point(color = colors$primary, size = 2, shape = 4) +
  scale_x_date(date_breaks = date_breaks, date_labels = date_labels) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  labs(
    title = "Cumulative FDA MAUDE Reports",
    subtitle = "Growing trend analysis",
    x = "Date",
    y = "Cumulative Number of Reports",
    caption = "Source: FDA MAUDE Database"
  )
```

```{=typst}
#set page(
  flipped: false,
)
```

# Product Problem Analysis

```{r compute-product-problems}
#| echo: false

# Extract and count individual product problems
product_problems_vector <- parse_problem_lists(data$product_problems)

product_problems <- tibble(problems = product_problems_vector) %>%
  filter(
    !is.na(problems) &
    problems != "" &
    !problems %in% PRODUCT_EXCLUSIONS
  ) %>%
  count(problems, sort = TRUE) %>%
  slice_head(n = 15)

top_3_problems <- product_problems %>%
  head(3) %>%
  pull(problems)

total_product_problems <- sum(product_problems$n)
```

## Top Product Problems

```{r product-problems-plot}
#| label: fig-product-problems
#| fig-cap: "Most frequently reported product problems"
#| fig-width: 8
#| fig-height: 8

ggplot(product_problems, aes(x = reorder(problems, n), y = n)) +
  geom_col(fill = colors$highlight, alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 15 Product Problems Reported",
    subtitle = paste0("(", format(date_min, "%B %Y"), 
                     " - ", format(date_max, "%B %Y"), ")"),
    x = "Product Problem",
    y = "Number of Reports",
    caption = "Source: FDA MAUDE Database"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

::: {.callout-warning}
## Product Issues

The most frequently reported problems are:

1. **`r top_3_problems[1]`** - `r product_problems$n[1]` reports (`r round(100 * product_problems$n[1] / total_reports, 1)`%)
2. **`r top_3_problems[2]`** - `r product_problems$n[2]` reports (`r round(100 * product_problems$n[2] / total_reports, 1)`%)
3. **`r top_3_problems[3]`** - `r product_problems$n[3]` reports (`r round(100 * product_problems$n[3] / total_reports, 1)`%)

These top three issues account for **`r sum(product_problems$n[1:3])`** reports (**`r round(100 * sum(product_problems$n[1:3]) / total_reports, 1)`%** of all reports).
:::

# Patient Problem Analysis

```{r compute-patient-problems}
#| echo: false

# Extract and count patient problems
patient_problems_vector <- parse_problem_lists(data$patient_problems)

patient_problems <- tibble(problems = patient_problems_vector) %>%
  filter(
    !is.na(problems) &
    problems != "" &
    !problems %in% PATIENT_EXCLUSIONS
  ) %>%
  count(problems, sort = TRUE) %>%
  filter(n > 1)

total_patient_problems <- sum(patient_problems$n)
```

## Patient Problems Overview

```{r patient-problems-plot}
#| label: fig-patient-problems
#| fig-cap: "Patient problems reported in FDA MAUDE events"
#| fig-width: 8
#| fig-height: 7

ggplot(patient_problems, aes(x = reorder(problems, n), y = n)) +
  geom_col(fill = colors$info, alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Patient Problems Reported",
    subtitle = "Clinical outcomes and adverse events",
    x = "Patient Problem",
    y = "Number of Reports",
    caption = "Source: FDA MAUDE Database"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

# Manufacturer Analysis

## Top Manufacturers

```{r compute-manufacturers}
#| echo: false

# Calculate manufacturer statistics
top_manufacturers <- data %>%
  count(manufacturer_std, sort = TRUE) %>%
  slice_head(n = 10)

top_5_manufacturers <- top_manufacturers %>%
  head(5)

top_manufacturer <- top_5_manufacturers$manufacturer_std[1]
top_manufacturer_count <- top_5_manufacturers$n[1]
top_manufacturer_pct <- round(100 * top_manufacturer_count / total_reports, 1)
```

```{r manufacturer-plot}
#| label: fig-manufacturers
#| fig-cap: "Top 10 manufacturers by report volume"
#| fig-width: 8
#| fig-height: 6.5

ggplot(top_manufacturers, aes(x = reorder(manufacturer_std, n), y = n)) +
  geom_col(fill = colors$secondary, alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 10 Manufacturers by Report Volume",
    x = "Manufacturer",
    y = "Number of Reports",
    caption = "Source: FDA MAUDE Database"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

The top manufacturer, **`r top_manufacturer`**, accounts for **`r top_manufacturer_count`** reports (**`r top_manufacturer_pct`%** of total).

```{r manufacturer-table}
#| label: tbl-manufacturers
#| tbl-cap: "Top 5 manufacturers with report statistics"

manufacturer_table <- top_5_manufacturers %>%
  mutate(
    Rank = row_number(),
    Manufacturer = manufacturer_std,
    Reports = format(n, big.mark = ","),
    `% of Total` = sprintf("%.2f%%", n / total_reports * 100)
  ) %>%
  select(Rank, Manufacturer, Reports, `% of Total`)

kable(manufacturer_table, align = c("c", "l", "r", "r"))
```

# Device Brand Analysis

## Top Device Brands

```{=typst}
#set page(
  flipped: true,
)
```

```{r compute-brands}
#| echo: false

# Calculate brand statistics
top_brands <- data %>%
  count(brand_std, sort = TRUE) %>%
  slice_head(n = 10)

top_5_brands <- top_brands %>%
  head(5)

top_brand <- top_5_brands$brand_std[1]
top_brand_count <- top_5_brands$n[1]
top_brand_pct <- round(100 * top_brand_count / total_reports, 1)
```

```{r brand-plot}
#| label: fig-brands
#| fig-cap: "Top 10 device brands by report volume"
#| fig-width: 10
#| fig-height: 6

ggplot(top_brands, aes(x = reorder(brand_std, n), y = n)) +
  geom_col(fill = colors$primary, alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Top 10 Device Brands by Report Volume",
    x = "Brand",
    y = "Number of Reports",
    caption = "Source: FDA MAUDE Database"
  ) +
  theme(axis.text.y = element_text(size = 8))
```

```{=typst}
#set page(
  flipped: false,
)
```

The top device brand, **`r top_brand`**, accounts for **`r top_brand_count`** reports (**`r top_brand_pct`%** of total).

```{r brand-table}
#| label: tbl-brands
#| tbl-cap: "Top 5 device brands with report statistics"

brand_table <- top_5_brands %>%
  mutate(
    Rank = row_number(),
    Brand = brand_std,
    Reports = format(n, big.mark = ","),
    `% of Total` = sprintf("%.2f%%", n / total_reports * 100)
  ) %>%
  select(Rank, Brand, Reports, `% of Total`)

kable(brand_table, align = c("c", "l", "r", "r"))
```

## Brand Temporal Trends

```{r brand-cumulative}
#| label: fig-brand-cumulative
#| fig-cap: "Cumulative adverse event reports by top 5 brands"
#| fig-width: 8
#| fig-height: 6

brand_cumulative <- data %>%
  filter(brand_std %in% top_5_brands$brand_std) %>%
  arrange(brand_std, date_received) %>%
  group_by(brand_std) %>%
  mutate(
    cumulative = row_number(),
    brand_label = str_to_title(map_chr(brand_std, shorten_brand_name))
  ) %>%
  ungroup()

# Calculate appropriate date breaks
date_range_years <- as.numeric(difftime(max(data$date_received), 
                                        min(data$date_received), 
                                        units = "days")) / 365.25

if(date_range_years > 2) {
  date_labels <- "%Y"
  date_breaks <- "1 year"
} else {
  date_labels <- "%b %Y"
  date_breaks <- "3 months"
}

ggplot(brand_cumulative, aes(x = date_received, y = cumulative, color = brand_label)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "BrBG") +
  scale_x_date(date_labels = date_labels, date_breaks = date_breaks) +
  scale_y_continuous(labels = comma, breaks = pretty_breaks()) +
  labs(
    title = "Cumulative Reports by Device Brand",
    x = "Date Received",
    y = "Cumulative Reports",
    color = "Brand Name",
    caption = "Source: FDA MAUDE Database"
  ) +
  theme(legend.position = "right")
```

# Technical Appendix

## Data Source

- **Database**: OpenFDA Medical Device Adverse Events
- **API Access**: [https://open.fda.gov/apis/device/event/](https://open.fda.gov/apis/device/event/)

## Analysis Tools

- **Python 3.x** with pandas and rapidfuzz for data processing
- **R** with tidyverse, ggplot2, and lubridate for visualization
- **Fuzzy Matching**: RapidFuzz library for text standardization
- **Visualization**: ggplot2 with custom base R theme
- **Report Generation**: Quarto for reproducible analytics

## Exclusion Criteria

**Patient Problems Excluded:**

```
No Code Available, No Known Impact, Symptoms or Conditions,
No Information, No Consequences, Insufficient Information
```

**Product Problems Excluded:**

```
Adverse Event Without Identified Device, No Apparent Adverse Event,
Appropriate Term/Code Not Available, Unknown, Insufficient Information
```

## Report Metadata

- **Generated**: 2025-11-19 14:04:42
- **Dataset Version**: 2025-09-16
- **Total Records Analyzed**: 119
- **Analysis Pipeline**: OpenFDA Medical Device Adverse Event Analysis v2.1
